Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-14 Manuscript/manuscript.tex
Updated conclusions

In this study, we employed the electrometer concept to demonstrate that the binding of \ce{Na^+} and \ce{Ca^2+} ions
to a POPC lipid bilayer can be accurately described within a classical MD simulation 
force field, provided that electronic polarization is implicitly included via 
the electronic continuum correction~\cite{leontyev11}.
While the structural details of a POPC lipid bilayer simulated with the newly developed ECC-lipids
model agree with experiments with an accuracy comparable to the other state of the art lipid models,
the model also reproduces the experimental lipid head group order parameter responses to
a cationic surfactant, NaCl and CaCl$_2$ concentrations. 
It thus represents a significant improvement over 
other available lipid models, which overestimate cation binding affinities~\cite{catte16}.  
The ECC-lipids model is build upon
the Lipid14 POPC model~\cite{dickson14} by scaling the partial charges by a factor of 0.8
and reducing the Van der Waals radii by a factor of 0.89 for the headgroup, glycerol backbone and carbonyl atoms.

The good agreement with experiments enables us to 
interpret NMR experiments with atomistic details using MD simulations with the ECC-lipid model.
In line with previous interpretations of experimental data \cite{hauser76,hauser78,herbette84,binder02},
Ca$^{2+}$ ions interact mainly with phosphate oxygens.
However, the stoichiometry of calcium binding is significantly more complicated than
the simple ternary complex model, used to interpret NMR data, where one calcium binds to two POPC molecules~\cite{altenbach84}.
While complexes with one calcium ion bound to two lipids are the most probable also in the
ECC-lipids model, complexes of one or three lipids per one calcium
were observed to relatively abundant and also almost equally likely. While the success of the simple ternary
compex model in fitting NMR data is understandable based on the simulation results,
it cannot capture the complex nature of calcium binding to phospholipid bilayers
oberved in the present simulation.

Accurate description of cation binding to POPC bilayer paves the way for
simulations of complex biochemical systems at cellular membranes with realistically described
electrostatic interactions. To this end, the compatibility of
the ECC-lipids model with existing models for proteins and other biological molecules should to be verified and, if necessary, 
further adjustments along the ECC concept in force fields of molecules interacting with the lipids need to be done.

>>>>

In this study, we employed the electrometer concept to demonstrate that the binding of \ce{Na^+} and \ce{Ca^2+} ions to a POPC lipid bilayer can be accurately described within a classical MD simulation force field, provided that electronic polarization is implicitly included via the electronic continuum correction~\cite{leontyev11}. While the structural details of a POPC lipid bilayer simulated with the newly developed ECC-lipids model agree with experiments with an accuracy comparable to the other state of the art lipid models, the model also reproduces the experimental lipid head group order parameter responses to a cationic surfactant, NaCl, and CaCl$_2$ concentrations. It thus represents a significant improvement over other available lipid models, which overestimate cation binding affinities~\cite{catte16}. The ECC-lipids model is built upon the Lipid14 POPC model~\cite{dickson14} by scaling the partial charges by a factor of 0.8 and reducing the Van der Waals radii by a factor of 0.89 for the headgroup, glycerol backbone, and carbonyl atoms.

The good agreement with experiments enables us to interpret NMR experiments with atomistic details using MD simulations with the ECC-lipid model. In line with previous interpretations of experimental data \cite{hauser76,hauser78,herbette84,binder02}, Ca$^{2+}$ ions interact mainly with phosphate oxygens. However, the stoichiometry of calcium binding is significantly more complicated than the simple ternary complex model, used to interpret NMR data, where one calcium binds to two POPC molecules~\cite{altenbach84}. While complexes with one calcium ion bound to two lipids are the most probable also in the ECC-lipids model, complexes of one or three lipids per one calcium were observed to relatively abundant and also almost equally likely. While the success of the simple ternary complex model in fitting NMR data is understandable based on the simulation results, it cannot capture the complex nature of calcium binding to phospholipid bilayers observed in the present simulation.

Accurate description of cation binding to POPC bilayer paves the way for simulations of complex biochemical systems at cellular membranes with realistically described electrostatic interactions. To this end, the compatibility of the ECC-lipids model with existing models for proteins and other biological molecules should be verified and, if necessary, further adjustments following ECC concept to the force fields of the interacting molecules with the lipids should be addressed.
